---
title: "Pop stats and genetic distances"
author: "Femke Batsleer"
date: "15-2-2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages and data

```{r packages, message=FALSE, warning=FALSE}
library("dplyr")
library("tidyr")
library("tidyverse")
library("adegenet")
library("adespatial")
library("spdep")
library("poppr")
library("hierfstat")
library("diveRsity")
library("graph4lg")
library("miscTools")
library("stringr")
library("RColorBrewer")
library("MASS")
library("reshape2")
library("foreign")
library("GGally")
library("ggplot2")
library("ggsci")
```


```{r data}
data_genind <- readRDS("../data/genind/genind_pop.RDS")
data_genind_popyear <- readRDS("../data/genind/genind_popyear.RDS")
metadatapop <- read.dbf("../data/GIS/Population_centroids_selection_collapsedyear.dbf")
metadatapopyear <- read.dbf("../data/GIS/Populations_centroids_selection_all_Lambert.dbf")
```


## General info loci

Summary, number of alleles per locus, sample size per pop, heterozygosity per locus

```{r loci, echo=FALSE}
#Basic info dataset
data_genind
#number of alleles per locus
as.data.frame(table(data_genind$loc.fac))
#number of alleles in total
as.data.frame(table(data_genind$loc.fac)) %>% summarise(total=sum(Freq), average=mean(Freq),
                                                        min=min(Freq), max=max(Freq))
#sample size for each site
freq <- summary(data_genind$pop)
(sample_sizes <- as.data.frame(freq, col.names=c("freq")) %>% 
  tibble::rownames_to_column("pop") %>% arrange(freq) %>%
    rmarkdown::paged_table())

#graph info
sum_all <- summary(data_genind)
par(mfrow=c(2,2))
plot(sum_all$n.by.pop, sum_all$pop.n.all, xlab="Colonies sample size",
     ylab="Number of alleles",main="Alleles numbers and sample sizes",
     type="n")
text(sum_all$n.by.pop,sum_all$pop.n.all,lab=names(sum_all$n.by.pop))
barplot(sum_all$loc.n.all, ylab="Number of alleles",
        main="Number of alleles per locus", las=2)
barplot(sum_all$Hexp-sum_all$Hobs, main="Heterozygosity: expected-observed",
        ylab="Hexp - Hobs", las=2)
barplot(sum_all$n.by.pop, main="Sample sizes per population",
        ylab="Number of genotypes",las=3)

basic.stats(data_genind)

#percentage of NA's
summary(as.data.frame(data_genind@tab) %>% pivot_longer(cols=1:ncol(.)))
nrow(as.data.frame(data_genind@tab) %>% pivot_longer(cols=1:ncol(.)))
```


## Table for basic info per population

```{r}
#coordinates per pop (centroid of all samples)
coordinates_pops <- metadatapop %>% dplyr::select(Population, MEAN_X, MEAN_Y)
#frequency per pop per year
freq <- as.data.frame(summary(data_genind_popyear$pop)) %>% rename(freq = `summary(data_genind_popyear$pop)`) %>% rownames_to_column("PopulationYear") %>%
   mutate(year = substr(PopulationYear, nchar(PopulationYear)-4+1, nchar(PopulationYear))) %>%
  mutate(Population = substr(PopulationYear, 1, nchar(PopulationYear)-4)) %>%
  mutate(year= as.numeric(year)) %>%
  dplyr::select(-PopulationYear)
#number_pop
number_pop <- metadatapop %>% dplyr::select(Population, Number_pop)
  


summary_popsyear <- metadatapopyear %>% dplyr::select(Region, Population, Year) %>% left_join(number_pop, by="Population") %>%
  dplyr::select(Region, Number_pop, Population, Year) %>%
  left_join(coordinates_pops, by="Population") %>%
  mutate(year = as.numeric(Year)) %>% dplyr::select(-Year) %>%
  left_join(freq, by=c("Population", "year")) %>%
  arrange(Number_pop, year)
summary_popsyear
# write.csv(summary_popsyear, file="summary_properties_perpopperyear.csv")



freq_collapsed <- as.data.frame(summary(data_genind$pop)) %>% rename(freq = `summary(data_genind$pop)`) %>% rownames_to_column("Population")

summary_pops <- metadatapop %>% dplyr::select(Region, Number_pop, Population, MEAN_X, MEAN_Y, years_samp) %>%
  left_join(freq_collapsed, by=c("Population")) %>%
  arrange(Number_pop)
summary_pops
# write.csv(summary_pops, file="summary_properties_perpop.csv")
```


## Basic population genetic measures per population

private alleles, allelic richness, heterozygosity, Fis
(observed heterozygosity also says something about diversity; difference between both about one of the assumptions of HW not met (small pop size, selection, Wahlund effect,...)).  
In haplodiploidy, one actually expects lower Ho (due to purging, but also drift is faster and have a smaller effective population size due to haplodiploidy; Lester & Selander 1979)  
Fis probably not reliable due to haplodiploidy: higher inbreeding expected anyway (but see Liautard & Sundström (2005) Insects Sociaux, for estimate of inbreeding for haplodiploids based on individual-based level, using relatedness-measures)  
  
  Allelic richness:  
  * lowest in Wetteren, Kortenhoeff, Kalmthout2, Harchies, Lagland  
  * highest at westcoast, slightly decreasing towards east coast  
  * further decreasing in Kempen  
  He-Ho:  
  * inbreeding high in midcoast (small populations that have inbreeding, although connected to rest of coast, see Fst below)  
  * inbreeding higher in some populations in Kempen  
  * inbreeding high in Wallonia  
  * inbreeding (surprisingly) low in Wetteren, Kortenhoeff, Kalmthout2 (Hamont-Achel), while these are quite differentiated (Fst) and have low allelic richness. Possibility: bottleneck (Moyra's thesis, Molecular Ecology book p109-110)?  



```{r}
#private alleles
private_alleles <- private_alleles(data_genind, count.alleles=F) %>% apply(MARGIN = 1, FUN = sum)
(private_alleles <- as.data.frame(private_alleles) %>%
  add_rownames("pop") %>% arrange(private_alleles) %>%
    rmarkdown::paged_table())
#doesn't say much at this scale probably (small scales; probably more informative when looking between regions)

#allelic richness
#Basic population statistics
#allelic richeness
AR <- as_tibble(allelic.richness(data_genind)$Ar) #for each locus separately
nrow(AR) #number of loci
#change with dplyr to mean and Sd for each location
AR_mean <- AR %>% summarise(across(everything(), mean, na.rm=TRUE))
AR_sd <- AR %>% summarise(across(everything(), sd, na.rm=TRUE))

AR_pop <- as_tibble(t(AR_mean %>% bind_rows(AR_sd)), rownames = "pop") %>%
  rename(mean=V1, sd=V2) %>% mutate(se = sd/sqrt(21))

AR_pop <- AR_pop %>% left_join(metadatapop, by=c("pop"="Population"))

#make a graph for allelic richness
plot_AR <- ggplot(AR_pop, aes(x=reorder(pop, Number_pop), y=mean)) +
  geom_point() +  geom_errorbar(aes(ymin=mean-se, ymax=mean+se)) +
  xlab("Population") + ylab("Rarefied Allelic Richness") +  ylim(c(1.5, 3.5)) +
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1))
plot_AR

#per region
(AR_region <- AR_pop %>% group_by(Region2) %>%
  summarize(meanAR = mean(mean), minAR=min(mean), maxAR=max(mean)))
t.test(mean~Region2, data=AR_pop)

#Heterozygosities
Ho <- as_tibble(basic.stats(data_genind)$Ho)
He <- as_tibble(basic.stats(data_genind)$Hs)
nrow(Ho)
Ho_mean <- Ho %>% summarise(across(everything(), mean, na.rm=TRUE))
Ho_sd <- Ho %>% summarise(across(everything(), sd, na.rm=TRUE))
He_mean <- He %>% summarise(across(everything(), mean, na.rm=TRUE))
He_sd <- He %>% summarise(across(everything(), sd, na.rm=TRUE))

Ho_pop <- as_tibble(t(Ho_mean %>% bind_rows(Ho_sd)), rownames = "pop") %>%
  rename(H_mean=V1, H_sd=V2) %>% mutate(H_se=H_sd/sqrt(21)) %>%
  add_column(H_measure = "Ho")
He_pop <- as_tibble(t(He_mean %>% bind_rows(He_sd)), rownames = "pop") %>%
  rename(H_mean=V1, H_sd=V2) %>% mutate(H_se=H_sd/sqrt(21)) %>%
  add_column(H_measure = "He")

(Ho_region <- Ho_pop %>% left_join(metadatapop, by=c("pop"="Population")) %>% group_by(Region2) %>%
  summarize(meanHo = mean(H_mean), minHo=min(H_mean), maxHo=max(H_mean)))
t.test(H_mean~Region2, data=Ho_pop %>% left_join(metadatapop, by=c("pop"="Population")))

(He_region <- He_pop %>% left_join(metadatapop, by=c("pop"="Population")) %>% group_by(Region2) %>%
  summarize(meanHe = mean(H_mean), minHo=min(H_mean), maxHo=max(H_mean)))
t.test(H_mean~Region2, data=He_pop %>% left_join(metadatapop, by=c("pop"="Population")))

H_pop_diff <- rename(Ho_pop, Ho_mean = H_mean, Ho_sd = H_sd, Ho_se=H_se, Ho_measure = H_measure) %>%
  left_join(He_pop, by=c("pop")) %>%
  mutate(He_Ho_mean = H_mean - Ho_mean, He_Ho_se = H_se + Ho_se) %>% left_join(metadatapop, by=c("pop"="Population"))

H_pop <- Ho_pop %>% bind_rows(He_pop) %>%
  mutate(pop=as.character(pop)) %>%
  left_join(metadatapop, by=c("pop"="Population"))
plot_H <- ggplot(H_pop, aes(x=reorder(pop, Number_pop), y=H_mean, fill=H_measure)) +
  geom_bar(stat = "identity", position="dodge") +
  geom_errorbar(aes(ymin=H_mean-H_se, ymax=H_mean+H_se), position="dodge") +
  xlab("Population") + ggtitle("Observed and expected heterozygosity") +
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1))
plot_H

plot_Hdiff <- ggplot(H_pop_diff, aes(x=reorder(pop, Number_pop), y=He_Ho_mean)) +
  geom_bar(stat="identity") +
  geom_errorbar(aes(ymin = He_Ho_mean-He_Ho_se, ymax = He_Ho_mean+He_Ho_se)) +
  xlab("Population") + ggtitle("Expected - observed heterozygosity") +
  theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))
plot_Hdiff
#watch out for effect of low sample sizes!


Fis <- as_tibble(basic.stats(data_genind)$Fis)
Fis_mean <- Fis %>% summarise(across(everything(), mean, na.rm=TRUE))
Fis_sd <- Fis %>% summarise(across(everything(), sd, na.rm=TRUE))

Fis_pop <- as_tibble(t(Fis_mean %>% bind_rows(Fis_sd)), rownames="pop") %>%
  rename(Fis_mean=V1, Fis_sd=V2) %>% mutate(Fis_se=Fis_sd/sqrt(21)) %>%
  left_join(metadatapop, by=c("pop"="Population"))

plot_Fis <- ggplot(Fis_pop, aes(x=reorder(pop, Number_pop), y=Fis_mean)) +
  geom_point() + geom_errorbar(aes(ymin=Fis_mean-Fis_se, ymax=Fis_mean+Fis_se)) +
  xlab("Population") + ylab(expression(F["IS"])) +
  theme(axis.text.x  = element_text(angle = 90, vjust=0.5, hjust=1)) +
  geom_hline(yintercept=0) + geom_hline(yintercept=mean(Fis_pop$Fis_mean), linetype="dashed")
plot_Fis
```


### Adding the population statistics to the table

```{r}
#Add all this info to the table above

summary_pops_stats <- summary_pops %>% full_join(private_alleles, by=c("Population"="pop"))  %>%#add private alleles
  full_join(AR_pop %>% rename("AR"="mean", "AR_sd"="sd", "AR_se"="se") %>% dplyr::select(pop, AR, AR_sd, AR_se), #add Allelic Richness
            by=c("Population"="pop")) %>%
  full_join(He_pop %>% rename("He_mean"="H_mean", "He_sd"="H_sd", "He_se"="H_se") %>%  #add expected heterozygosity
              dplyr::select(pop, He_mean, He_sd, He_se),
            by=c("Population"="pop")) %>%
  full_join(Ho_pop %>% rename("Ho_mean"="H_mean", "Ho_sd"="H_sd", "Ho_se"="H_se") %>%  #add observed heterozygosity
              dplyr::select(pop, Ho_mean, Ho_sd, Ho_se),
            by=c("Population"="pop")) %>%
  full_join(Fis_pop %>% dplyr::select(pop, Fis_mean, Fis_sd, Fis_se), #add inbreeding coefficient
            by=c("Population"="pop"))

# write.csv(summary_pops_stats, "summary_pop_statistics.csv")

```


## Nei's genetic distance

poppr function nei.dist or hierfstat (but is dat Nei's distance?)
hierfstat::genet.dist: Ds (Nei's standard genetic distance)
confidence intervals? Toch Genalex or other stand alone program?


```{r Neis genetic distance, fig.height=10, fig.width=12}
nei.dist_m <- as.data.frame(as.matrix(hierfstat::genet.dist(data_genind, method="Ds")))
colnames(nei.dist_m) <- popNames(data_genind) #add popnames to columns
rownames(nei.dist_m) <- popNames(data_genind) #add popnames to rows

metadatapop_assign <- metadatapop %>% mutate(pop = gsub('\\ ', "", Population)) #to be able to join data

nei.dist_df <- nei.dist_m %>% rownames_to_column(var="pop1") %>%
  pivot_longer(cols = Simliduinen:(ncol(nei.dist_m)+1), names_to="pop2", values_to="actual") %>% #put the values from matrix -> long format
  left_join(metadatapop_assign %>% #adding the number of the pops
                dplyr::select(Population, Number_pop) %>% rename("X_1"="Number_pop"),
              by=c("pop1"="Population")) %>%
    left_join(metadatapop_assign %>%
                dplyr::select(Population, Number_pop) %>% rename("X_2"="Number_pop"), 
              by=c("pop2"="Population"))

write.csv(nei.dist_df, "Neisgeneticdistance_hierfstat.csv")

#plot Nei's distance

Ds_tile <-  ggplot(nei.dist_df, aes(x=reorder(as.character(X_1), X_1), y=reorder(as.character(X_2), X_2), fill=actual)) + #x=reorder(pop1, X_1), y=reorder(pop2, X_2)
  geom_tile() +
  xlab("Population ID") + ylab("Population ID") +
  scale_fill_distiller(palette = "PuBuGn", direction=1, name=expression(D[S])) +#RdYlBu
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        text = element_text(size = 17))
Ds_tile

#average 
(meanDs_pop <- nei.dist_df %>%
    dplyr::filter(pop1!=pop2) %>%
    group_by(pop1) %>% summarise(Ds=mean(actual), sd=sd(actual)) %>% arrange(desc(Ds)) %>%
    left_join(dplyr::select(metadatapop_assign, Population, Number_pop, Region), by=c("pop1"="Population")) %>%
    rmarkdown::paged_table())

# write.csv(meanDs_pop, "average Nei's genetic distance per pop.csv")
```

## Pairwise Fst values

Calculated with package diveRsity. The first chunk of code has to be run to actually calculate the indices (should be uncommented to run), but takes a lot of time to run (because of the bootstraps).

```{r fig.height=10, fig.width=12}
#to work with package diveRsity: convert genind to genepop-file
#https://rdrr.io/github/romunov/zvau/man/writeGenPop.html

# genpop <- genind_to_genepop(data_genind, "all_genind_genepop.txt")
# 
# # #
# gendata <- readGenepop(infile = "all_genind_genepop.txt", gp = 3, bootstrap = F)
# diffs <- fastDivPart(infile = "all_genind_genepop.txt", outfile = "./differentation diveRsity/13092021", pairwise=T)
# # 
# fst <- diffCalc(infile = "all_genind_genepop.txt", outfile="diffs",
#                 fst=T, pairwise=T, bs_pairwise=T, boots = 999)
# write.table(fst$bs_pairwise$Fst, "Outputs/Fst_diveRsity1000boots.txt")
# write.table(fst$bs_pairwise$D, "Outputs/D_diveRsity1000boots.txt")
# write.table(fst$bs_pairwise$gst, "Outputs/gst_diveRsity1000boots.txt")
# write.table(fst$bs_pairwise$Gst, "Outputs/Gst_diveRsity1000boots.txt")
# write.table(fst$bs_pairwise$GGst, "Outputs/GGst_diveRsity1000boots.txt")

#define function to read differentation measure from above files
read_diffmeasure <- function(file = "Outputs/Fst_diveRsity1000boots.txt"){
  diff_df <-  read.table(file, head=T) %>%
  separate(populations, c("pop1", "pop2"), sep="\\s*vs\\s*") %>%
  mutate(pop1 = gsub("*_.*", "", pop1)) %>%
  mutate(pop2 = gsub("*_.*", "", pop2))
  #add all all data bind_rows, but with pop1 and pop2 swapped
  #remove all duplicates?
  diff_df_all <- diff_df %>% dplyr::select(pop2, pop1, actual, lower, upper) %>%
    rename("pop1"="pop2", "pop2"="pop1") %>%
    bind_rows(diff_df)
  #join with metadata to get right order in pop1 and pop2
  diff_df_all <- diff_df_all %>%
    left_join(metadatapop_assign %>%
                dplyr::select(pop, Number_pop) %>% rename("X_1"="Number_pop"),
              by=c("pop1"="pop")) %>%
    left_join(metadatapop_assign %>%
                dplyr::select(pop, Number_pop) %>% rename("X_2"="Number_pop"), 
              by=c("pop2"="pop"))
  return(diff_df_all)
}

fst_df_all <- read_diffmeasure("Outputs/Fst_diveRsity1000boots.txt")

fst_tile <-  ggplot(fst_df_all, aes(x=reorder(pop1, X_1), y=reorder(pop2, X_2), fill=actual)) +
  geom_tile() +
  scale_fill_distiller(palette = "BrBG") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
fst_tile


#leave out non-sign fst-values
fst_df_sign <- fst_df_all %>% mutate(Fst = ifelse(lower < 0 & upper > 0, NA, actual))

fst_tile_sign <-  ggplot(fst_df_sign, aes(x=reorder(as.character(X_1), X_1), y=reorder(as.character(X_2), X_2), fill=Fst)) + #x=reorder(pop1, X_1), y=reorder(pop2, X_2)
  geom_tile() +
  xlab("Population ID") + ylab("Population ID") +
  scale_fill_distiller(palette = "PuBuGn", direction=1, name=expression(F[ST])) +#RdYlBu
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        text = element_text(size = 17))
fst_tile_sign

#table to summarise mean pairwise D per pop
(meanfst_pop <- fst_df_all %>% group_by(pop1) %>% summarise(Fstmean=mean(actual), sd=sd(actual)) %>% arrange(desc(Fstmean)) %>%
    left_join(dplyr::select(metadatapop_assign, pop, Number_pop, Region), by=c("pop1"="pop")) %>%
    rmarkdown::paged_table())
# write.csv(meanfst_pop, "average pairwise Fst per pop.csv")

####Add regions as second level of axes: https://stackoverflow.com/questions/20571306/multi-row-x-axis-labels-in-ggplot-line-chart
```


## Alternatives for Fst: Djost, Gst


Djost, calculated above (Jost 2008, good for highly polymorphic markers); Gst (+ G'st, G''st are possible but not given)  

Hierfstat is another package which might have other possibilities (but not used here).  


```{r diff correlations, fig.height=10, fig.width=12}
# read measures
D_df_all <- read_diffmeasure("Outputs/D_diveRsity1000boots.txt")
Gst_df_all <- read_diffmeasure("Outputs/gst_diveRsity1000boots.txt")

#bind measures into one dataframe

diffmeasures <- rename(fst_df_all, Fst=actual, Fst_lower=lower, Fst_upper=upper) %>%
  left_join(rename(D_df_all, Djost=actual, Djost_lower=lower, Djost_upper=upper), by=c("pop1", "pop2")) %>%
  left_join(rename(Gst_df_all, Gst=actual, Gst_lower=lower, Gst_upper=upper), by=c("pop1", "pop2"))

ggpairs(dplyr::select(diffmeasures, Fst, Djost, Gst))
```

### Plot Djost

```{r plot D, fig.height=10, fig.width=12}
#leave out non-sign values
D_df_sign <- D_df_all %>% mutate(Djost = ifelse(lower < 0 & upper > 0, NA, actual))


D_tile_sign <-  ggplot(D_df_sign, aes(x=reorder(as.character(X_1), X_1), y=reorder(as.character(X_2), X_2), fill=Djost)) +
  geom_tile() +
  xlab("Population ID") + ylab("Population ID") +
  scale_fill_distiller(palette = "PuBuGn", direction=1, name="D") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        text = element_text(size = 17))
D_tile_sign



#table to summarise mean pairwise D per pop
(meanD_pop <- D_df_all %>% group_by(pop1) %>% summarise(Djost=mean(actual), sd=sd(actual)) %>% arrange(desc(Djost)) %>%
    left_join(dplyr::select(metadatapop_assign, pop, Number_pop, Region), by=c("pop1"="pop")) %>%
    rmarkdown::paged_table())
# write.csv(meanD_pop, "average pairwise D per pop.csv")

#tables for average D's etc

```


### plot Gst

```{r plot Gst, fig.height=10, fig.width=12}
#leave out non-sign values
Gst_df_sign <- Gst_df_all %>% mutate(Gst = ifelse(lower < 0 & upper > 0, NA, actual))

Gst_tile_sign <-  ggplot(Gst_df_sign, aes(x=reorder(pop2, X_2), y=reorder(pop1, X_1), fill=Gst)) +
  geom_tile() +
  xlab("pop2") + ylab("pop1") +
  scale_fill_distiller(palette = "BrBG") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
Gst_tile_sign

#table to summarise mean pairwise Fst per pop
(meanGst_pop <- Gst_df_all %>% group_by(pop1) %>% summarise(Gst=mean(actual), sd=sd(actual)) %>% arrange(desc(Gst)) %>%
    rmarkdown::paged_table())
```



