---
title: "Isolation By Distance"
author: "Femke Batsleer"
date: "15-2-2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages and data

```{r packages, message=FALSE, warning=FALSE}
library("dplyr")
library("tidyr")
library("tidyverse")
library("adegenet")
library("adespatial")
library("spdep")
library("poppr")
library("hierfstat")
library("diveRsity")
library("graph4lg")
library("miscTools")
library("stringr")
library("RColorBrewer")
library("MASS")
library("reshape2")
library("foreign")
library("scales")
library("car")
```

## Loading data

data from France and Wallonia is left out, as these more distant pops are not sampled on a level of whole landscape-coverage. IBD says less when not most of in-between pops are representatively sampled.

```{r data}
data_genind <- readRDS("../data/genind/genind_pop.RDS") %>%
  popsub(exclude=c("Sangatte", "Slack", "Mt St Frieux", "Harchies","Lagland"))
metadatapop <- read.dbf("../data/GIS/Population_centroids_selection_collapsedyear.dbf") %>%
  filter(!Population %in% c("Sangatte", "Slack", "Mt St Frieux", "Harchies","Lagland"))
```


## Plots for isolation by distance (IBD)

good to read through: https://adegenet.r-forge.r-project.org/files/tutorial-basics.pdf  


```{r}
#with adegenet
#make sure Dgen and Dgeo are in the same order
metadatapop_order <- data.frame(genorder = popNames(data_genind)) %>%
  left_join(metadatapop, by=c("genorder"="Population")) %>%
  mutate(genorder = gsub('\\ ', "", genorder))

nrow(metadatapop_order) == length(popNames(data_genind))

#coordinates populations and distance calculations
coords_df <- dplyr::select(metadatapop_order, genorder, MEAN_X, MEAN_Y) %>%
  column_to_rownames("genorder")
Dgeo <- dist(coords_df, method="euclidian") #distances in matrix format
# 
# distances in dataframe format
Dgeo_df <- melt(as.matrix(Dgeo, labels=T), varnames = c("row", "col")) %>%
  rename('value_geo' = 'value') %>%
  mutate(row= as.character(row), col = as.character(col))  %>%
  mutate(row = gsub('\\ ', "", row), col = gsub('\\ ', "", col))

##Neis genetic distance
Nei_df <- read.csv("../3 - Pop stats and gen dist/Neisgeneticdistance_hierfstat.csv") %>%
  mutate(pop1 = gsub("\\ ", "", pop1), pop2 = gsub("\\ ", "", pop2))

Dnei_df <- Dgeo_df %>% left_join(Nei_df %>% dplyr::select(-X_1, -X_2, -X), by=c("row"= "pop1", "col"="pop2")) %>%
  filter(value_geo >0)#delete diagonal values

# add/join regions etc to be able to plot separately
Dfst_numbers <- Dnei_df %>% left_join(metadatapop_order, by=c("row"="genorder")) %>% #add and rename first of pairwise region
  rename("row_npop" = "Number_pop") %>%
  left_join(metadatapop_order, by=c("col"="genorder")) %>% #add and rename second of pairwise region
  rename("col_npop" = "Number_pop") %>%
  #give names to pairwise values (inland=pairwise within inland region, coastal=pairwise within coast & between=pairwise between coast and inland)
  mutate(region = ifelse(row_npop > 40 & col_npop > 40, "inland",
                                              ifelse(row_npop <= 40 & col_npop <= 40, "coastal", "between")))

#rearrange order of region to get inland (less points) on top of coastal
coastalinland_df <- Dfst_numbers[order(as.numeric(factor(Dfst_numbers$region))),] %>%
  dplyr::filter(region!="between")

ibd_coastalinland <- ggplot(coastalinland_df,
                            aes(x=value_geo/1000, y=actual, col=region)) +
  geom_point() +
  ylab(expression(D[S])) + xlab("pairwise dist (km)") +
  geom_smooth(method = "lm", se = F) +
  scale_color_manual(values = c("coastal" = "#dfc27d", "inland" = "#80cdc1")) +
  theme_bw()
ibd_coastalinland
```


## Permutation tests to check if slopes and intercepts are different

```{r}
library(lmPerm)
#dataset to be used
permdata <- coastalinland_df %>% dplyr::select(value_geo, actual, Region2.x, Region.x, Region2.y, Region.y) %>%
  mutate(dist = value_geo/1000) #%>% #distance in m
  #mutate(dist_resc = rescale(dist, to=c(0,1)), actual_resc = rescale(actual, to=c(0,1)))

perm.model <- lmp(formula = actual~dist*Region2.x, data=permdata, perm="Prob", maxIter=10000, check_contrasts=F)
summary(perm.model)
Anova(perm.model, type=3)
```


## RDA analysis

Ideas from following:
https://www.scielo.br/j/gmb/a/kRQX63dHrC9m7xVnSxCfVqd/?format=pdf&lang=en : here they select PCA-axes according to broken-stick criterion. We use all axes. With selection, there would be inflation of R2 compared to real R2 & then you are only able to use relatively.  

Other ideas/options:  
** https://popgen.nescent.org/2018-03-27_RDA_GEA.html #not for distance (environmental variables only)

** https://bio-protocol.org/bio101/r10202477 : links pairwise Fst to geographical distances: PCNM to transform matrix of geographical distance into vectors

** slightly different option: supporting info in Meirmans (2015) https://onlinelibrary.wiley.com/doi/10.1111/mec.13243 for r-script: links allele frequencies to polynomials of spatial coordinates.  

### RDA on complete dataset

```{r}
library(vegan)

#first, get distances (geographic/euclidian distances and genetic distances) into matrix-format
##geographic distances
geodist.df <-  Dgeo_df %>% left_join(Nei_df %>% dplyr::select(-X_1, -X_2, -X), by=c("row"= "pop1", "col"="pop2")) %>%
  mutate(value_geo = value_geo/1000) %>%
  dplyr::select(row, col, value_geo) %>% arrange(row, col) %>%
  pivot_wider(names_from = col, values_from = value_geo) %>%
  column_to_rownames(var = "row")
geodist.m <- as.dist(geodist.df, diag=T)
##genetic distances
gendist.df <- Dgeo_df %>% left_join(Nei_df %>% dplyr::select(-X_1, -X_2, -X), by=c("row"= "pop1", "col"="pop2")) %>%
  dplyr::select(row, col, actual) %>% arrange(row, col) %>%
  pivot_wider(names_from = col, values_from = actual) %>%
  column_to_rownames(var = "row")
gendist.m <- as.dist(gendist.df, diag=T)

###RDA on full dataset####
#dbRDA: PCoA on genetic distances, link this with spatial coordinates in rda
pcoa_nei <- wcmdscale(gendist.m, eig=T)

#retain all axes
eig_resp <- pcoa_nei$points

#put them in dataframe together to get focal coordinates correct
coords_df_pop <- coords_df %>% rownames_to_column(var = "Population")
data_rda <- data.frame(eig_resp) %>% rownames_to_column(var = "Population") %>% full_join(coords_df_pop, by="Population")
nrow(data_rda) == nrow(coords_df_pop)
eig_resp <- dplyr::select(data_rda, Dim1:Dim9)
coords_var <- dplyr::select(data_rda, MEAN_X, MEAN_Y)

#perform RDA with response eigenvectors from PCoA on Nei's genetic distance and explanatory variables coordinates
nei_rda <- rda(eig_resp ~ ., data= coords_var)
nei_rda$tot.chi
nei_test <- anova.cca(nei_rda, step=1000)
nei_test

RsquareAdj(nei_rda)
```


### RDA on coast and inland separately


```{r}
#get nei-matrix for coast and inland
##coast
coast_pop <- metadatapop_order %>% filter(Region == 'Flanders-coast') %>% dplyr::select(genorder)
coast_pop.v <- as.vector(coast_pop$genorder)

geodist_coast.df <- Dgeo_df %>% left_join(Nei_df %>% dplyr::select(-X_1, -X_2), by=c("row"= "pop1", "col"="pop2")) %>%
  mutate(value_geo = value_geo/1000) %>%
  filter(row %in% coast_pop.v) %>% filter(col %in% coast_pop.v) %>%
  dplyr::select(row, col, value_geo) %>% arrange(row, col) %>%
  pivot_wider(names_from = col, values_from = value_geo) %>%
  column_to_rownames(var = "row")
geodist_coast.m <- as.dist(geodist_coast.df, diag=T)

gendist_coast.df <- Dgeo_df %>% left_join(Nei_df %>% dplyr::select(-X_1, -X_2), by=c("row"= "pop1", "col"="pop2")) %>%
  filter(row %in% coast_pop.v) %>% filter(col %in% coast_pop.v) %>%
  dplyr::select(row, col, actual) %>% arrange(row, col) %>%
  pivot_wider(names_from = col, values_from = actual) %>%
  column_to_rownames(var = "row")
gendist_coast.m <- as.dist(gendist_coast.df, diag=T)

##inland
inland_pop <- metadatapop_order %>% filter(Region == 'Flanders-inland') %>% dplyr::select(genorder)
inland_pop.v <- as.vector(inland_pop$genorder)

geodist_inland.df <- Dgeo_df %>% left_join(Nei_df %>% dplyr::select(-X_1, -X_2), by=c("row"= "pop1", "col"="pop2")) %>%
  mutate(value_geo = value_geo/1000) %>%
  filter(row %in% inland_pop.v) %>% filter(col %in% inland_pop.v) %>%
  dplyr::select(row, col, value_geo) %>% arrange(row, col) %>%
  pivot_wider(names_from = col, values_from = value_geo) %>%
  column_to_rownames(var = "row")
geodist_inland.m <- as.dist(geodist_inland.df, diag=T)

gendist_inland.df <- Dgeo_df %>% left_join(Nei_df %>% dplyr::select(-X_1, -X_2), by=c("row"= "pop1", "col"="pop2")) %>%
  filter(row %in% inland_pop.v) %>% filter(col %in% inland_pop.v) %>%
  dplyr::select(row, col, actual) %>% arrange(row, col) %>%
  pivot_wider(names_from = col, values_from = actual) %>%
  column_to_rownames(var = "row")
gendist_inland.m <- as.dist(gendist_inland.df, diag=T)

#consistency checks
colnames(geodist_coast.df) == colnames(gendist_coast.df)
rownames(geodist_coast.df) == rownames(gendist_coast.df)
colnames(geodist_coast.df) == rownames(geodist_coast.df)

#PCoA
pcoa_nei_coast <- wcmdscale(gendist_coast.m, eig=T)
pcoa_nei_inland <- wcmdscale(gendist_inland.m, eig=T)

#retain all axes
eig_resp_coast <- pcoa_nei_coast$points
eig_resp_inland <- pcoa_nei_inland$points

#put them in dataframe together to get focal coordinates correct
data_rda_coast <- data.frame(eig_resp_coast) %>% rownames_to_column(var = "Population") %>% left_join(coords_df_pop, by="Population")
eig_resp_coast <- dplyr::select(data_rda_coast, Dim1:Dim6)
coords_var_coast <- dplyr::select(data_rda_coast, MEAN_X, MEAN_Y)

data_rda_inland <- data.frame(eig_resp_inland) %>% rownames_to_column(var = "Population") %>% left_join(coords_df_pop, by="Population")
eig_resp_inland <- dplyr::select(data_rda_inland, Dim1:Dim3)
coords_var_inland <- dplyr::select(data_rda_inland, MEAN_X, MEAN_Y)

#perform RDA with response eigenvectors from PCoA on nei and explanatory variables coordinates
nei_rda_coast <- rda(eig_resp_coast ~ ., data= coords_var_coast)
nei_rda_coast
nei_rda_coast$tot.chi
nei_test_coast <- anova.cca(nei_rda_coast, step=1000)
nei_test_coast

nei_rda_inland <- rda(eig_resp_inland ~ ., data= coords_var_inland)
nei_rda_inland
nei_rda_inland$tot.chi
nei_test_inland <- anova.cca(nei_rda_inland, step=1000)
nei_test_inland

RsquareAdj(nei_rda_coast)
RsquareAdj(nei_rda_inland)

plot(nei_rda_coast)
plot(nei_rda_inland)
```



## Mantel tests

Just to check (see main manuscript and references therein for comments on Mantel tests).  
Results are similar as with RDA.  

```{r}
#All populations
all_mantel <- mantel.randtest(geodist.m, gendist.m, nrepet = 9999)
all_mantel
plot(all_mantel)

#Coastal populations
#mantel test
coast_mantel <- mantel.randtest(geodist_coast.m, gendist_coast.m, nrepet = 9999)
coast_mantel
plot(coast_mantel)

#Inland populations
#mantel test
inland_mantel <- mantel.randtest(geodist_inland.m, gendist_inland.m, nrepet = 9999)
inland_mantel
plot(inland_mantel)
```